A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-26 reviews/response.Rmd

> I hold a similar sentiment regarding the calendar plot in Figure 12. It's a great plot but I don't necessary think this is a plot that show cased the benefit of tsibble.

Yes, the calendar plot just requires a data frame. We've added the forecasting example to back up this plot: how the weather information improves the demand forecasts.

> Figure 11 with missing data was a great demonstration for tsibble and the data times series pipeline shown in Figure 3. It may be helpful to reduce the opacity of the points as there is a lot of overplotting. The top line which lumps a number of customers together is not particularly informative here. Perhaps it is better replaced with a histogram.

Thanks. We separated the remaining customers to the bottom panel with points of some transparency. We have kept the line range plot for the consistency between two panels.

@@

> Authors may like to add/change `spread`/`gather` verbs to `pivot_longer`/`pivot_wider` with the recent move to deprecate the former in tidyr. Likewise authors may like to consider to replace “long data” to “longer data” and “wide data” to “wider data” as long/wide implies absolute terms. <https://tidyr.tidyverse.org/dev/articles/pivot.html>

We'd like to upgrade `spread`/`gather` verbs to `pivot_longer`/`pivot_wider`. But they are not S3 generics yet, which delays the implementation. Conceptually the newly melted columns will be added to the key specification.

> Manuscript does implicitly feel more time series data oriented rather than a general temporal data. E.g. references of time points as index seems inherent from other time series data infrastructure (i.e. zoo and xts) and the presented case studies is not representative of any longitudinal study. This does not mean it should be changed but index to me was not as intuitive to be the temporal variable.

Time indexes lay the contextual foundation to both time series and longitudinal data. `zoo` and `xts` store indices as metadata due the matrix structure, but tsibble's indices are explicitly a data column that can be manipulated. The index may be a date-time variable, as is usual in time series, but could be as simple as an ordered integer as is common in longitudinal data. It is also customizable to any variable that can be ordered. We have made this clearer in Section 3.1 where index specification is explained.

@@

Tsibble is aimed at general temporal data, to unify **data** structures underlying time series and longitudinal problems. In Section 4 of the revised manuscript, we clarify the different focuses of time series and longitudinal **analysis** and modelling. However, we have kept the emphasis on time series for the examples, because there are more functions available for time series model building on tsibble.

> Figure 4 animation doesn't play so authors may like to ensure that the final publication works. Github version works.

Thanks for pointing it out. We include a gif version of the animation as supplementary materials.

### 2019-08-27 reviews/response.Rmd

---
title: Response to reviewers
author: JCGS-19-044
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

We thank the reviewers for their helpful feedback which has led to many improvements in the paper. The point by point description of changes are below: the reviewers' comments are in red and our response is in black.

This manuscript has undergone an overhaul to address reviewer's comments:

1. Section 2.2 (previously 2.3) has been extended to include reviews of other software than R (Stata and SAS) time series and longitudinal data standards.
2. We have rewritten Section 3 "Contextual semantics" to give a definition of tsibble in relation to tidy data, and we expand on the four rules of a tsibble in details.
3. We have rewritten Section 4 "Temporal data pipelines" to describe a big picture about each module with the focus of temporal data instead of time series before. We have added Section 4.4 to summarise what a fluent pipeline looks like for a tsibble.
4. We have removed the mosaic plot along with the paragraph from Section 6.1, and added the forecasting example to the smart meter data in Section 6.2 to show tsibble's capability.

### 2019-08-27 tsibble.Rmd
section 2.1-2

## Time series and longitudinal data {#sec:time-series}

Temporal data problems are typically grouped into two types of analysis, time series and longitudinal. Despite being exactly the same data input, the representation of time series and longitudinal data diverges due to different modeling focuses.

Time series can be univariate or multivariate, and for modeling require relatively long lengths (i.e., large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. Most statistical software represent such time series as vectors or matrices. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wider format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric vectors. Time indices are implicitly inferred as attributes or meta-information. The number of observations must be the same across all the series. Data wrangling, from the form that data arrives in to this specialist format, can be mysterious and frustrating. This kind of format is model-centric, and the diagram in the style of Figure \@ref(fig:tsibble-fit) would place the "model" at the center of the analytical universe, and all the transformations and visualizations would hinge on that format. This is contrary to the **tidyverse** conceptualization, which holistically captures the full data workflow.

For longitudinal analysis, researchers and analysts are primarily interested in explaining trends across and variations among individuals, and making inference about a broader population. Longitudinal data or panel data typically assumes fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs that measurements for individuals are taken at different time points. Thus, the primary format required for modeling such data is stacked data, blocks of measurements for each individual, with columns indicating panels, times of measurement and the measurements themselves. (The tidy data name for this would be **longer format**.) Evidently, this data organization saves storage space for many sparse cells, compared to structuring it into the wider format which would have missing values in many cells. An appealing feature is that data is structured in a semantic manner with reference to observations and variables, with panel and time stated explicitly. It is conducive to examining the data in many different ways and leading to more comprehensive exploration.

## Existing time series and longitudinal data standards

In R [@R-base], time series and longitudinal data are of different representations. The native `ts` object and the enhancement of **zoo** [@zoo2005] and **xts** [@R-xts], assemble time series into wider matrices with implicit time indices. A relatively new R package **tibbletime** [@R-tibbletime] proposed a data class of *time tibble* to represent time series in heterogeneous longer format. It only requires an index variable to be declared. However, as proposed in Section \@ref(sec:semantics) a more rigid data structure is required for temporal analytics and models. The **plm** [@plm2008] and **panelr** [@R-panelr] packages both manage longitudinal data in terms of longer format.

Stata [@Stata] provides two commands, `tsset` and `xtset`, to declare time series and panel data respectively, both of which need specifying panel and time variables. Note they are in fact the same command, according to the manual. The underlying data arrangement is the unifying longer form. Either time series or panel commands are sufficient to apply to both types. The SAS software [@SAS-STAT] handles them in a similar way as Stata.

@@

@@

The authors would like to thank Mitchell O'Hara-Wild for many discussions on the software development and Davis Vaughan for contributing ideas on rolling window functions. We also thank Stuart Lee for the feedback on this manuscript. We are grateful for anonymous reviewers for helpful feedback that has led to many improvements of the paper. This article was created with knitr [@knitr] and R Markdown [@rmarkdown]. The project's Github repository <https://github.com/earowang/paper-tsibble> houses all materials required to reproduce this article and a history of the changes.

>>>>

## Comparing time series and longitudinal data {#sec:time-series}

Temporal data problems often fall into two types of analysis, time series and longitudinal. Both of these may have very similar data input, but the representation for modelling is typically different. Time series analysis tends to focus on the dependency with series, and the cross-correlation between series. Longitudinal analysis tends to focus on overall temporal patterns across demographic or experimental treatment strata, that incorporates within subject dependency.

Time series can be univariate or multivariate, and require relatively long lengths (i.e., large $T$) for modeling. With this large $T$ property, the series can be handled as stochastic processes for the primary purpose of forecasting, and characterizing temporal dynamics. Due to an expectation of regular time intervals, and equal length across series, multivariate time series are typically assumed to be in the format where each column contains a single time series, and time is specified implicitly. This also implies that data are columns of homogeneous types: either all numeric or all non-numeric. To wrestle data, in whatever format that it arrives in, to this modeling format can be frustrating. The format could be considered to be model-centric, rather than data-centric, and thus throws the analyst into the deep end of the pool, rather than allowing them to gently wade to the modeling stage from the shallow end. The expectation is that the "model" is at the center of the analytical universe. This is contrary to the **tidyverse** conceptualization (Figure \@ref(fig:tsibble-fit)), which holistically captures the full data workflow. More support needs to be provided, in the form of consistent tools and data structures, to transform the data into the analytical cycle.

Longitudinal data (or panel data) typically assumes fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs that measurements for individuals are taken at different time points, and in different quantities. The primary format required for modeling is stacked data, blocks of measurements for each individual, with columns indicating panels, times of measurement and the measurements themselves.   An appealing feature is that data is structured in a semantic manner with reference to observations and variables, with panel and time explicitly stated.

## Existing data standards

In R [@R-base], time series and longitudinal data are of different representations. The native `ts` object and the enhancement of **zoo** [@zoo2005] and **xts** [@R-xts], assemble time series into wide matrices with implicit time indexes. A relatively new R package **tibbletime** [@R-tibbletime] proposed a data class of *time tibble* to represent time series in heterogeneous long format. It only requires an index variable to be declared. However, this is insufficient, and a more rigid data structure is required for temporal analytics and modeling. The **plm** [@plm2008] and **panelr** [@R-panelr] packages both manage longitudinal data in long format.

Stata [@Stata] provides two commands, `tsset` and `xtset`, to declare time series and panel data respectively, both of which require explicit panel and time index specification.  The underlying data arrangement is only long form, for both types of data. The functions pay no attention to whether the data is for time series analysis or longitudinal data analysis. The SAS software [@SAS-STAT] also handles both types of data in the same way as Stata.

@@

 (The tidy data name for this would be **wider format**.)

@@

The authors would like to thank Mitchell O'Hara-Wild for many discussions on the software development and Davis Vaughan for contributing ideas on rolling window functions. We also thank Stuart Lee for the feedback on this manuscript. We are grateful for anonymous reviewers for helpful feedback that has led to many improvements in the paper. This article was created with knitr [@knitr] and R Markdown [@rmarkdown]. The project's Github repository <https://github.com/earowang/paper-tsibble> houses all materials required to reproduce this article and a history of the changes.
